Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-05-01 TEBenchmarkPaper/article.tex
(Semi-)final fiddling

Other than $\alpha$-equivalence and equation symmetry, we specifically
\emph{ignore} logical relationships between syntactically distinct statements,
such as one equation being implied by another. Whilst logically sound,
second-guessing the ground truth in this way would harm other aspects which
influence interestingness (for example, a more general statement is more widely
applicable, but it might also be harder to comprehend).

@@

Each sample is only explored once by each tool, so that we cover as many
independent samples as possible to better estimate how a tool's performance
generalises to unseen inputs. How we combine these data into an aggregate
summary depends on what we are interested in measuring.

One general question we might ask is how a tool's performance scales with
respect to the input size (the number of given definitions). This is
straightforward to measure by varying the sample size, but we need some way to
combine the results from samples of the same size.

@@

    \node (in)  [io]{Theorem Proving Benchmark};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

@@

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this (arbitrary)
threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
a few days we chose a timeout of 300 seconds, since that covered most of the
successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
diminishing returns. During analysis, duplicate samples (caused by our sampling
restrictions) were found and discarded, so only 14 samples of size 1 were used
and 30 of size 2.

@@

We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
since it satisfies the desiderata of $\S$\ref{section:prep}: each benchmark
problem has standalone type and function definitions, making their separation
trivial; known examples from the software verification and inductive theorem
proving literature are included, ensuring relevance to those fields; the format
includes the higher-order functions and inductive datatypes we are interested
in; it is large enough to pose a challenge to current MTE tools; plus it is
accompanied by tooling to convert its custom format (an extension of
SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
Isabelle.

@@

We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{section:mte}. In order to thoroughly benchmark QuickSpec, we need
to automate some of the decisions which are normally left up to the user:

@@

The time taken to explore samples of different sizes is shown in
figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
and these occurred for all sample sizes but are more common for larger samples
(over half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 133 conjectures from a
sample of size 19.

Precision and recall are shown in figure~\ref{figure:quickspec_precRec}.
Since QuickSpec generates monotonically more conjectures as definitions are
added to a signature (assuming sufficient testing), its decreasing precision
can't be due to finding fewer wanted conjectures at larger sizes. This is
supported by the relative flatness of the recall results. Rather, the number of
conjectures generated is increasing at a higher rate than the size of the ground
truth. These extra conjectures may involve those ``padding'' definitions in a
sample which don't contribute to its ground truth, or may be ``uninteresting''
relationships between the dependencies of different ground truth properties.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered down to a more desirable sub-set by another post-processing step.
Alternatively, rather than exploring all of the given definitions together,
multiple smaller signatures could be selected from the input by predicting which
combinations are likely to lead to interesting conjectures; this would avoid
both the ``padding'' and the cross-dependency relationships. Both of these
methods could improve the precision, although care would be needed to avoid a
large reduction in recall. The latter option could also improve the running
time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
signatures may be faster to explore than a single large one. Such improvements
would also need to be ``generic'', to avoid over-fitting to this particular
benchmark.

@@

The most striking result is how rapidly IsaCoSy's running time, shown in
figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
size 6 failed, with most timing out and a few aborting early due to running out
of memory. Like in our preliminary testing, this increase appears exponential,
so even large increases to the timeout would not produce many more successful
observations. A significant amount of IsaCoSy's time (around 50 seconds) was
spent loading the generated theory (containing all distinct datatypes and
functions from TIP); this overhead is unfortunate, but since it is constant
across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of IsaCoSy's output, shown in
figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
proportions that IsaCoSy is capable of discovering interesting conjectures, and
the graphs appear to follow the same shapes as those of QuickSpec (decreasing
precision, flat recall), although the error margins are too wide to be
definitive.

@@

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which is compared against
IsaCoSy and HipSpec is compared using precision/recall
in~\cite{claessen2013automating}. The name comes from its embedding in the
Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
can be used to generate new concepts and conjectures. Variables within a scheme
are instantiated automatically and this drives the invention process. For
example, the concept of ``repetition'' can be encoded as a scheme, and
instantiated with existing encodings of zero, successor and addition to produce
a definition of multiplication. The same scheme can be instantiated with this
new multiplication function to produce exponentiation.

IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
described in more detail in $\S$\ref{sec:existing-tools}, since these are the
tools we chose to evaluate and compare to demonstrate our benchmarking
methodology. QuickSpec has since evolved to version 2~\cite{smallbone2017quick},
which replaces the distinct enumeration and testing steps with a single,
iterative algorithm similar to that of IsaCoSy. Generated conjectures are fed
into a Knuth-Bendix completion algorithm to form a corresponding set of rewrite
rules. As expressions are enumerated, they are simplified using these rules and
discarded if equal to a known expression. If not, QuickCheck tests whether the
new expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

@@

Another notable MTE implementation, distinct from those based in Isabelle and
Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions)~\cite{roy}. The project's goal is to help
mathematicians explore the consequences of a set of axioms or a particular
concept definition by deducing, via the chosen logic, a set of theorems which
they entail. Like other MTE approaches, one of its main challenges is to
automatically evaluate interestingness: to distinguish important theorems from
results which, although they follow from a set of axioms, are of little
mathematical interest. Evaluation of these theorems was performed by the
system's developer, which highlights the need for a general method of evaluation
and comparison for these tools.

Whilst there are many reasonably objective benchmarks for mathematical tasks
such as automated theorem proving, the precision/recall analysis shown
in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
the only quantitative comparison of these recent tools we have found in the
literature. Our work is essentially an extension of this approach, to a larger
and more diverse set of examples. The suitability of the TIP theorem proving
benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
since its developers are also those of the IsaCoSy and QuickSpec tools we have
tested. This goes some way to ensuring that our demonstration is a faithful
representation of the task these tools were designed for, whilst our independent
repurposing of the problem set reduces the risk that the benchmark is
tailor-made for these tools; or, conversely, that the tools over-fit to these
particular problems.

>>>>

We specifically \emph{ignore} other logical relationships between syntactically
distinct statements, such as one equation being implied by another. Whilst
logically sound, second-guessing the ground truth in this way would harm other
aspects which influence interestingness (for example, a more general statement
is more widely applicable, but it might also be harder to comprehend).

@@

Each sample is only explored once by each tool, so that we cover as many
independent samples as possible to better estimate how a tool's performance
generalises to unseen inputs. How we combine these data into an aggregate
summary depends on what we are interested in measuring. One general question we
might ask is how a tool's performance scales with respect to the input size (the
number of given definitions). This is straightforward to measure by varying the
sample size, but we need some way to combine the results from samples of the
same size.

@@

    \node (in)  [io]{Theorem Proving Benchmark};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

@@

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this (arbitrary)
threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
a few days we chose a timeout of 300 seconds, since that covered most of the
successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
diminishing returns. During analysis, duplicate samples (caused by our
requirement that samples have a non-empty ground truth) were found and
discarded, so only 14 samples of size 1 were used and 30 of size 2.

@@

We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
problem has standalone type and function definitions, making their separation
trivial; known examples from the software verification and inductive theorem
proving literature are included, ensuring relevance to those fields; the format
includes the higher-order functions and inductive datatypes we are interested
in; it is large enough to pose a challenge to current MTE tools; plus it is
accompanied by tooling to convert its custom format (an extension of
SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
Isabelle.

@@

We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
need to automate some of the decisions which are normally left up to the user:

@@

The time taken to explore samples of different sizes is shown in
Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
failures were due to timing out, and these occurred for all sample sizes but are
more common for larger samples (over half the runs for sample sizes 13 and 20
timed out). Runs which succeeded mostly finished within 30 seconds, generating
few or no conjectures; those taking longer generated more conjectures, with the
most being 133 conjectures from a sample of size 19.

Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
Since QuickSpec generates monotonically more conjectures as definitions are
added to a signature (assuming sufficient testing), its decreasing precision
can't be due to finding fewer wanted conjectures at larger sizes. This is
supported by the relative flatness of the recall results. Rather, the number of
conjectures generated is increasing at a higher rate than the size of the ground
truth. These extra conjectures may involve those ``padding'' definitions in a
sample which don't contribute to its ground truth, or may be ``uninteresting''
relationships between the dependencies of different ground truth properties.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered down to a more desirable sub-set by another post-processing step.
Alternatively, rather than exploring all of the given definitions together,
multiple smaller signatures could be selected from the input by predicting which
combinations are likely to lead to interesting conjectures; this would avoid
both the ``padding'' and the cross-dependency relationships. Both of these
methods could improve the precision, although care would be needed to avoid a
large reduction in recall. The latter option could also improve the running
time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
signatures may be faster to explore than a single large one. Such improvements
would also need to be ``generic'', to avoid over-fitting to this particular
benchmark.

@@

The most striking result is how rapidly IsaCoSy's running time, shown in
Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
size 6 failed, with most timing out and a few aborting early due to running out
of memory. Like in our preliminary testing, this increase appears exponential,
so even large increases to the timeout would not produce many more successful
observations. A significant amount of IsaCoSy's time (around 50 seconds) was
spent loading the generated theory (containing all distinct datatypes and
functions from TIP); this overhead is unfortunate, but since it is constant
across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of IsaCoSy's output, shown in
Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
proportions that IsaCoSy is capable of discovering interesting conjectures, and
the graphs appear to follow the same shapes as those of QuickSpec (decreasing
precision, flat recall), although the error margins are too wide to be
definitive.

@@

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which has also been
quantitatively compared against IsaCoSy and HipSpec using precision/recall
analysis~\cite{claessen2013automating}. The name comes from its embedding in the
Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
can be used to generate new concepts and conjectures. Variables within a scheme
are instantiated automatically and this drives the invention process. For
example, the concept of ``repetition'' can be encoded as a scheme, and
instantiated with existing encodings of zero, successor and addition to produce
a definition of multiplication. The same scheme can be instantiated with this
new multiplication function to produce exponentiation.

IsaCoSy and QuickSpec (the conjecture generation component of HipSpec) are
described in more detail in $\S$\ref{sec:existing-tools}, since these are the
tools we chose to evaluate and compare for $\S$\ref{sec:application}. QuickSpec
has since evolved to version 2~\cite{smallbone2017quick}, which replaces the
distinct enumeration and testing steps with a single, iterative algorithm
similar to that of IsaCoSy. Generated conjectures are fed into a Knuth-Bendix
completion algorithm to form a corresponding set of rewrite rules. As
expressions are enumerated, they are simplified using these rules and discarded
if equal to a known expression. If not, QuickCheck tests whether the new
expression can be distinguished from the known expressions through random
testing: those which can are added to the set of known expressions. Those which
cannot be distinguished are conjectured to be equal, and the rewrite rules are
updated.

@@

Another notable MTE implementation, distinct from those based in Isabelle and
Haskell, is the MATHsAiD project (Mechanically Ascertaining Theorems from
Hypotheses, Axioms and Definitions)~\cite{roy}. Unlike the tools above, which
generate \emph{conjectures} that may later be sent to automated provers,
MATHsAiD directly generates \emph{theorems}, by making logically valid
inferences from a given set of axioms and definitions. Evaluation of the
interestingness of these theorems was performed qualitatively by the system's
developer, which highlights how these tools could benefit from the availability
of an objective, repeatable, quantitative method of evaluation and comparison
such as ours.

Whilst there are many reasonably objective benchmarks for mathematical tasks
such as automated theorem proving, the precision/recall analysis shown
in~\cite{claessen2013automating}, and described further in $\S$\ref{sec:te}, is
the only quantitative comparison of these recent MTE tools we have found in the
literature. Our work is essentially an extension of this approach, to a larger
and more diverse set of examples. The suitability of the TIP theorem proving
benchmark for our purposes, detailed in $\S$\ref{sec:tip}, is not coincidental,
since its developers are also those of the IsaCoSy and QuickSpec tools we have
tested. This goes some way to ensuring that our demonstration is a faithful
representation of the task these tools were intended to solve; our independent
repurposing of this problem set, in a way it was not designed for, reduces the
risk that the benchmark is tailor-made for these tools (or that the tools
over-fit to these particular problems).
